\textbf{Wavelet Scattering Transforms Applied to Whale Vocalisations}

### 2024-05-18 dissertation/content/p1_intro.tex
started intro

The most popular \ac{tf} analysis is the \ac{stft}.

In terms of equation (\ref{eqn:gentf}), a \ac{stft} is computed with 
\begin{equation}
    \label{eqn:stftfilter}
    \upsilon_k[n] =  w[n] e^{j \frac{2\pi k n}{N}},
\end{equation}
where $w$ is a window function defined by the user. Various \acp{fft} are use to compute this transform, so the properties of $w$ is typically specified by the \ac{fft} length, and the window type (Hann, Hamming, Blackman, etc.). The hop-length (or downsample factor) $d$ is an additional parameter specified by the user.

@@

Suppose that we design $\Upsilon$ to contain $Q$ filters per octave. A frequency change by a factor $a$ shifts the frequency index of the fundamental to $(Q\log_2 (a) + 1) k_0$. Similarly, the $b$'th harmonic's frequency index is shifted to $(Q\log_2 (a) + 1) (Q\log_2 (b) + 1) k_0$. For example, if $a=2$, the fundamental's frequency index is shifted to $k_0 + Q$ (as opposed to $2 k_0$ for the \ac{stft}). This property is referred to a frequency-shift invariance, since small frequency deformations of the order $Q$ do not significantly change the frequency index structure of the \ac{tf} decomposition. The \ac{stft} is much more sensitive to such deformations across its entire structure for harmonically rich signals.

@@

$uppsi$ can the be scaled by a factor $\lambda$ such that
\begin{equation}
    \upsilon_k(t) = \lambda_k \uppsi(\lambda_k t).
\end{equation}

>>>>

The most popular \ac{tf} analysis is the \ac{stft}. In terms of equation (\ref{eqn:gentf}), a \ac{stft} is computed with 
\begin{equation}
    \label{eqn:stftfilter}
    \upsilon_k[n] =  w[n] e^{j \frac{2\pi k n}{N}},
\end{equation}
where $w$ is a window function defined by the user. Various \acp{fft} are use to compute this transform, so the properties of $w$ is typically specified by the \ac{fft} length, and the window type (Hann, Hamming, Blackman, etc.). The hop-length (or downsample factor) $d$ is an additional parameter specified by the user. It is possible to choose $w$ and $d$ such that the \ac{stft} becomes an invertible operator -- the original signal can be entirely recovered.

@@

Suppose that we design $\Upsilon$ to contain $Q$ filters per octave. A frequency change by a factor $a$ shifts the frequency index of the fundamental to $(Q\log_2 (a) + 1) k_0$. Similarly, the $b$'th harmonic's frequency index is shifted to $(Q\log_2 (a) + 1) (Q\log_2 (b) + 1) k_0$. For example, if $a=2$, the fundamental's frequency index is shifted to $k_0 + Q$ (as opposed to $2 k_0$ for the \ac{stft}). This property is referred to a frequency-shift invariance, since small frequency deformations of the order $Q^{-1}$ do not significantly change the frequency index structure of the \ac{tf} decomposition. The \ac{stft} is much more sensitive to such deformations across its entire structure for harmonically rich signals.

@@

$\uppsi$ can the be scaled by a factor $\lambda$ such that
\begin{equation}
    \upsilon_k(t) = \lambda_k \uppsi(\lambda_k t).
\end{equation}

### 2024-05-18 dissertation/content/p2_intro.tex
started intro

We can construct a wavelet filterbank which, similar to MFCCs, introduces time and frequency-shift invariance in a similar manner to the Mel spectrogram. This process is constructed in the opposite order (frequency-shift, then time-shift invariance) to the Mel spectrogram. However, its properties remain similar \cite{ws}.

@@

The notation in equations (\ref*{eqn:general_scalogram}) and (\ref*{eqn:general_scattering}) conforms the general \ac{tf} description of chapter \ref*{chap:p1i}, although the specific wavelet scattering operators in chapters \ref*{chap:p2} to \ref*{chap:p3} follow a different standard. These operators are known as wavelet scattering, which formalise the notation of ``averaging a \ac{tf}-decomposition'' in terms of digital signal processing. Chapters \ref*{chap:p2} to \ref*{chap:p3} discuss the specifics, flavours and implementation of scattering operators at length.

>>>>

We can construct a wavelet filterbank which introduces time and frequency-shift invariance in a similar manner to the Mel spectrogram. This process is constructed in the opposite order (frequency-shift, then time-shift invariance) to the Mel spectrogram. However, its properties remain similar \cite{ws}.

@@

The notation in equations (\ref*{eqn:general_scalogram}) and (\ref*{eqn:general_scattering}) conforms the general \ac{tf} description of chapter \ref*{chap:p1i}, although the specific wavelet scattering operators in chapters \ref*{chap:p2} to \ref*{chap:p3} follow a different standard. These operators are known as wavelet scattering, which formalise the notion of ``averaging a \ac{tf}-decomposition'' in terms of signal processing theory. Chapters \ref*{chap:p2} to \ref*{chap:p3} discuss the specifics, flavours and implementation of scattering operators at length.

### 2024-05-18 dissertation/content/p3_intro.tex
started intro

\section{Types of Scattering Transforms}
Existing scattering transforms are listed and described below for convenience:
\begin{enumerate}
    \item 1D scattering \citep{1dscattering1, ws}. This is the first-introduced scattering transform, which serves as a generalisation of MFCCs. It improves on MFCCs by recovering the information lost due to invariance filter (averaging) operations. Morlet wavelets are used, with cascaded scattering levels having its own number of filter per octave.
    \item 2D scattering \citep{2dscattering}. Gabor filters modified to 0-mean, are used to construct the filterbank. The filterbank consists of exponentially placed filters in one axis, which is then rotated by equally spaced angles to cover the frequency plane. This filterbank only has a single filter per octave ($Q=1$). In particular, in polar coordinates $(r, \theta)$ in the frequency domain, filters are placed exponentially along $r$, but linearly along $\theta$.
    \item Harmonic scattering \citep{harmonicscattering} (3 dimensions). The spherical harmonics are used as filters which consitute an orthogonal basis in 3 dimensions. These filters have ``lobes'' that extend in a multitude of dimensions, depending on the harmonic, in the frequency domain.
    \item 3D scattering -- extension of 2D scattering \citep{3dscattering}. The filterbank is constructed similar to 2D scattering. In spherical coordinates $(\rho, \theta, \phi)$ in the frequency domain, filters are spaced exponentially along $\rho$, and linearly along $\theta$ and $\phi$.
    \item Joint time-frequency \citep{ws_joint_tf,jointtfscattering2} scattering -- 2D filters on 1D scattering scalogram. A first level of scattering in performed in 1D to produce a scalogram ($X[k, m]$). 2D scattering is then performed on subsequent levels, with $X[k, m]$ as input. The 2D wavelets are separable - a product of 1D wavelets. This allows 1D scattering on audio to have features that are also dependent on adjascent frequency bins. It is conceptually similar to a \ac{cnn} operating on a \ac{stft} spectrogram. In the frequency domain with axes $(\omega_1, \omega_2)$, the 2D filters are constructed as the cross-product between filter sets that placed exponentially in both $\omega_1$ and $\omega_2$.
    This is the only literature considering separable filters.
\end{enumerate}

@@

>>>>

\section{Types of Scattering Transforms}
Existing scattering transforms are listed and described below for convenience:
\begin{enumerate}
    \item 1D scattering \citep{1dscattering1, ws}. This is the first-introduced scattering transform, which serves as a generalisation of MFCCs. It improves on MFCCs by recovering the information lost due to invariance filter (averaging) operations. Morlet wavelets are used, with cascaded scattering levels having its own number of filter per octave.
    \item 2D scattering \citep{2dscattering}. Gabor filters modified to 0-mean, are used to construct the filterbank. The filterbank consists of exponentially placed filters in one axis, which is then rotated by equally spaced angles to cover the frequency plane. In particular, in polar coordinates $(r, \theta)$ in the frequency domain, filters are placed exponentially along $r$ with a single filter per octave, but linearly along $\theta$.
    \item Harmonic scattering \citep{harmonicscattering} (3 dimensions). The spherical harmonics are used as filters which consitute an orthogonal basis in 3 dimensions. These filters have ``lobes'' that extend in a multitude of dimensions, depending on the harmonic, in the frequency domain.
    \item 3D scattering -- extension of 2D scattering \citep{3dscattering}. The filterbank is constructed similar to 2D scattering. In spherical coordinates $(\rho, \theta, \phi)$ in the frequency domain, filters are spaced exponentially along $\rho$ with a single filter per octave, and linearly along $\theta$ and $\phi$.
    \item Joint time-frequency \citep{ws_joint_tf,jointtfscattering2} scattering -- 2D filters on 1D scattering scalogram. A first level of scattering in performed in 1D to produce a scalogram ($X[k, m]$). 2D scattering is then performed on subsequent levels, with $X[k, m]$ as input. The 2D wavelets are separable - a product of 1D wavelets. This allows 1D scattering on audio to have features that are also dependent on adjascent frequency bins. It is conceptually similar to a \ac{cnn} operating on a \ac{stft} spectrogram. In the frequency domain with axes $(\omega_1, \omega_2)$, the 2D filters are constructed as the cross-product between filter sets that placed exponentially in both $\omega_1$ and $\omega_2$.
    This is the only literature considering separable filters.
\end{enumerate}

@@

Recently, wavelet scattering operations implemented as \acp{cnn}, referred to as a scattering network, have been of increasing interest in order to increase the effectiveness of \ac{cnn} filters. Results indicate that making some aspects of scattering networks learnable yields significant improvements compared to non-learnable filters. Particularly, fixed spatial and/or time filters with learnable across-channel filters seem to obtain the best results.

For gradient descent, it may be beneficial to perform convolutions directly, and not in the Fourier domain. Separable filters have been shown to be effective in a \ac{cnn} settings, while also allowing for much quicker convolution operations. Separable filters also allow for the independance of each dimension, which can be particularly useful for optimisations such as downsampling. This type of scheme is illustrated by higher-dimensional dyadic wavelet transforms such as the \ac{dwt}.

Downsampling schemes are trivial when the axes are independant -- a filterbank can be configured independantly for each axis and all operations for following axes do not dependent on previous filtering operations. When a mix of time and spatial dimensions are present, techniques such as 2D, 3D and harmonic wavelet scattering may not be suitable, since their invariance filters cannot provide a certain amount of invariance for each dimension independently. Additionally, 2D and 3D scattering tradionally allow the user to create a filterbank with more than 1 filter per octave across $r$ and $\rho$ respectively.
